Calibrating seed-based alignment heuristics with Sesame

### 2020-02-22 manuscript.tex
Investigate the true vs best location

TODO: mention that this is MEM seeds only. What happens for other seed
types by the way?

Early in the development of the theory, we made the assumption that the
true location is alo the best. This hypothesis was required to claim that
a read is mapped to the true location whenever seeding is on target.
Indeed, if the target is in the candidate set, the exact nature of the
alignment algorithms ensures that the read will be mapped correctly. This
assumption is thus critical because it says that all mapping errors are
caused by off-target seeding.

To know how reasonable this statement is, we have to evaluate how often
the read is mapped incorrectly when seeding is on target. More accurately,
we have to consider the probability that the candidate set contains the
target plus a candidate that is more similar to the read than the target.

The key insight is to realize that MEM seeds tend to exclude the target
from the candidate set if it is not the best location. To develop an
intuition as to why this is the case, consider a read with one sequencing
error. If this is not the best location, then there exists a candidate
location that is a perfect match for the read. But in that case, the read
is covered by a single MEM seed and the target is masked so it cannot be
in the candidate set.

Using simulations, we can show that off-target seeding is much more
frequent than mapping errors after on-target seeding (Table~XXX). In other
words, for the vast majority of mapping errors, the target is not seeded.
This means that, the initial assumption that the true location is always
the best, although incorrect, does not appreciably modify the accuracy of
the estimates for MEM seeds.

For exact and skip-seeds, the conclusions are different.

>>>>

Early in the development of the theory, we distinguished the true location
from the best location, but we assumed that they are identical in order to
eliminate some practical considerations that would otherwise clutter the
exposition. Here we have presented a framework to compute the probability
that the \emph{true} location is in the candidate set. We have not said
anything about the probability that the best location is in the candidate
set, so our results do not help address the best location problem. Now it
remains to see whether they help address the true location problem.

The issue at hand is that the candidates are tested with an alignment
algorithm that returns the best location. So even if the target is in the
candidate set, the read may be mapped to another location. The question is
thus how often this happens. If this case is rare, our results give a good
approximation of the mapping errors (where errors are failures to map the
reads to their true location). Otherwise the estimates may not be so
useful in practice.

The answer depends on the seed type. We start with MEM seeds because they
will allow us to better explain what happens. The key insight is that MEM
seeds tend to exclude the target from the candidate set if it is not the
best location. To develop an intuition as to why this is the case,
consider a read with a single sequencing error. We claim that the target
was never in the candidate set if the read is mapped to an incorrect
location (\textit{i.e.} not the true location).

The argument goes as follows: if the target were in the candidate set,
then there would exist a candidate that is a perfect match for the read.
But then the read would be covered by a single MEM seed, the target would
be masked and it could not be in the candidate set, leading to a
contradiction.

\begin{table}
% MEM seeds.
\renewcommand{\arraystretch}{1.1}
\centering
\begin{tabular}{ccccc}
k & N & Prob. seeding error & Prob. other error & \% seeding \\
\hline
50  &  1  & $7.8 \cdot 10^{-4}$ & $9.5 \cdot 10^{-5}$ & $89.1\%$ \\
50  & 10  & $7.8 \cdot 10^{-3}$ & $9.1 \cdot 10^{-4}$ & $89.6\%$ \\
50  & 100 & $7.6 \cdot 10^{-2}$ & $5.7 \cdot 10^{-3}$ & $93.0\%$ \\
100 &  1  & $2.1 \cdot 10^{-5}$ & $7.6 \cdot 10^{-6}$ & $73.7\%$ \\
100 & 10  & $2.6 \cdot 10^{-4}$ & $7.1 \cdot 10^{-5}$ & $78.1\%$ \\
100 & 100 & $5.3 \cdot 10^{-3}$ & $5.9 \cdot 10^{-4}$ & $90.1\%$
\end{tabular}
\caption{My favorite table. $p = 0.01$ and $\mu = 0.06$.}
\label{table_mem}
\end{table}

In short, either seeding is on target and the read is mapped to the true
location, or seeding is off target (and the read is of course mapped
somewhere else). When there are more than one errors, the strict
equivalence breaks, but the trend remains the same. We can show this
using simulations for realistic sequencing conditions
(Table~\ref{table_mem}). For MEM seeds, the probability that the read is
not mapped to the true location is approximately the same as the
probability that the seeding is off target.

For exact and skip-seeds, the conclusions are different. The situation is
even the opposite if we consider reads with a single sequencing error. In
this case the target is always in the candidate set (if the read size is
greater than $2\gamma$, where $\gamma$ is the minimum seed length), so
mapping errors are never due to failures of the seeding heuristic.
Instead, mapping errors happen only when the true location is not the
best.

The trend can reverse for a high number of errors (most of the errors may
be due to seeding errors), but because these reads are rare, the mapping
errors are in general not due to failures of the seeding heuristic.

\begin{table}
% Exact seeds.
\renewcommand{\arraystretch}{1.1}
\centering
\begin{tabular}{ccccc}
k & N & Prob. seeding error & Prob. other error & \% seeding \\
\hline
50  &  1  & $3.3 \cdot 10^{-5}$  & $5.1 \cdot 10^{-4}$ & $6.0\%$ \\
50  & 10  & $2.9 \cdot 10^{-4}$  & $5.0 \cdot 10^{-3}$ & $5.4\%$ \\
50  & 100 & $1.0 \cdot 10^{-3}$  & $4.6 \cdot 10^{-2}$ & $2.2\%$ \\
100 &  1  & $1.5 \cdot 10^{-10}$ & $2.3 \cdot 10^{-5}$ & $< 0.1\%$ \\
100 & 10  & $1.2 \cdot 10^{-9}$  & $2.3 \cdot 10^{-4}$ & $< 0.1\%$ \\
100 & 100 & $2.6 \cdot 10^{-9}$  & $2.3 \cdot 10^{-3}$ & $< 0.1\%$
\end{tabular}
\caption{My least favorite table.}
\label{table_exact}
\end{table}

\begin{table}
% Skip-9 seeds.
\renewcommand{\arraystretch}{1.1}
\centering
\begin{tabular}{ccccc}
k & N & Prob. seeding error & Prob. other error & \% seeding \\
\hline
50  &  1  & $1.3 \cdot 10^{-4}$ & $4.9 \cdot 10^{-4}$ & $20.1\%$ \\
50  & 10  & $1.2 \cdot 10^{-3}$ & $4.8 \cdot 10^{-3}$ & $19.7\%$ \\
50  & 100 & $5.5 \cdot 10^{-3}$ & $4.5 \cdot 10^{-2}$ & $10.9\%$ \\
100 &  1  & $2.0 \cdot 10^{-8}$ & $2.3 \cdot 10^{-5}$ & $< 0.1\%$ \\
100 & 10  & $1.5 \cdot 10^{-7}$ & $2.3 \cdot 10^{-4}$ & $< 0.1\%$ \\
100 & 100 & $3.9 \cdot 10^{-7}$ & $2.3 \cdot 10^{-3}$ & $< 0.1\%$
\end{tabular}
\caption{My not so favorite table.}
\label{table_skip}
\end{table}

Tables~\ref{table_exact} and \ref{table_skip} shows the result of numeric
simulations for exact and skip-9 seeds, respectively. It appears that in
these cases, the probability that probability of mapping the read to an
incorrect location (\textit{i.e.} not the true location) is substantially
higher than the probability that seeding is off target. As a consequence,
seeding probabilities are not very useful to address true mapping problem
when exact and skip seeds are used. In this case, it is more accurate to
use the probabilities that the true location is not the best.
